---
title: "chapter6"
author: "Otso Aro"
date: "12/6/2019"
output: html_document
---

install.packages("dplyr", repos = "http://cran.us.r-project.org")
install.packages("ggplo2", repos = "http://cran.us.r-project.org" )
install.packages("GGally", repos = "http://cran.us.r-project.org")

library(dplyr)
library(ggplot2)
library(GGally)

# Exercise 6

## Reading tables into file and factor the categorical variables again
```{r}
BPRS <- read.table("data/BPRS")
RATS <- read.table("data/RATS")
str(BPRS)
str(RATS)

# Factor treatment & subject
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

# Factor variables ID and Group
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
```

## MABS 8 analysis for the RATS-data
```{r}
# Draw the plot
ggplot(RATS, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS$Weight), max(RATS$Weight)))
```
### Unstandadized plot of the RATS-plot

*asdas*

## Standadizing the RATS weight variable
```{r}
RATS <- RATS %>%
  group_by(Time) %>%
  mutate(stdWeight = ((Weight - mean(Weight)) / sd(Weight))) %>%
  ungroup()

# Plot again with the standardised bprs
ggplot(RATS, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")
```

## Creating a summary Graph for RATS (with standadized weight)
```{r}
# Number of WD, baseline (week 0) included
n <- max(RATS$Time)
n

# Summary data with mean and standard error of Weight by weekday 
RATSL <- RATS %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = (sd(Weight)/ sqrt(n) )) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot the mean profiles
ggplot(RATSL, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

## MABS analysis
```{r}

# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline time = 1).
RATS8S <- RATS %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATS8S)
RATS$Weight

# Draw a boxplot of the mean versus treatment
ggplot(RATS8S, aes(x = Group, y = mean)) +
  geom_boxplot() + ggtitle("Boxplot of weights in groups") +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), weekdays 1-64")

```
*Here we can see that all of the groups have an outlier. With group 1 and 3 the outlier is dragging the mean of the groups weights down, while in group 2 it is the opposite.*


## MABS 9 analysis for the BPRS-data

```{r}
str(BPRS)
# Plot the BPRS - data
ggplot(BPRS, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = subject)) + scale_x_continuous(name = "Week", breaks = seq(0, 8, 1))
```

```{r}
# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data=BPRS)
summary(BPRS_reg)
```
*Week seems to be a significant indicator in values of the phychiatric scale. This means that every week in treatments seems to significantly lower the persons mental illness scale.*

```{r}
install.packages("lme4", repos = "http://cran.us.r-project.org")
library(lme4)
# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | week), data = BPRS, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```



## Creating another random intercept and random slope model and plotting it
```{r}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + week * subject, data = BPRS, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of RATSL with the observed Weight values
ggplot(BPRS, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = subject)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Observed phyciatric scale") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to RATSL
BPRS <- BPRS %>% mutate(Fitted)

# draw the plot of RATSL with the Fitted values of weight
ggplot(BPRS, aes(x = week, y = Fitted, group = subject)) +
  geom_line(aes(linetype = subject)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "Fitted psychiatric scale values") +
  theme(legend.position = "top")
```








